---
title: "Practical Machine Learning Prediction Assignment"
date: "26 February 2016"
output: html_document
---
# Practical Machine Learning Prediction Assignment

## Background
Using data from Human Activity Recognition devices we are required to undertake data analysis on the data at http://groupware.les.inf.puc-rio.br/har.  
  
The goal of the excercise is to predict the manner in which exercise was conducted either correctly being classe "A" or the incorrect options being classe's "B through E".

``` {r}
suppressMessages(library("dplyr"))
require(dplyr)
suppressMessages(library("tree"))
require(tree)
suppressMessages(library("caret"))
require(caret)
require(ggplot2)
suppressMessages(library("rattle"))
require(rattle)
set.seed(5436)
```


## Getting the data
```{r, echo=TRUE}
if(!getwd()== "C:/R/pml"){
        dir.create("C:/R/pml");
        setwd("C:/R/pml")
        download.file("https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv", "pml_train.csv")
download.file("https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv", "pml_test.csv")
                        }
```


## Cleaning the data  
After undertaking an initial review of the data I remove factor variables and non numeric data from the initial view of the training data set being columns 1,2,5,6,7.

I then removed variables with:  
1. zero variance,  
2. columnns with NA values, and   
3. highly correlated data.


```{r, cache=TRUE, echo=TRUE}
options(warn = -1)
# Load test data 
train <- read.csv("pml_train.csv", na.strings = c("NA","","#DIV/0"))

# subset the train data to undertake validation prior to using the test data
inTrain = createDataPartition(train$classe  , p = 3/4)[[1]]

training <- train[inTrain,]
vldtn <- train[-inTrain,]

test <- read.csv("pml_test.csv", na.strings = c("NA","","#DIV/0"))


##  Cleaning the data
#Initially I retrieved only the information pertaining to accelerometers on the belt, forearm, arm, and dumbell.  This was a dumb move.
# subs <- grep("class|belt|forearm|arm|dumbell ", names(train), value = TRUE)
#training <- training[,subs]
training <- training[,-c(1,2,5,6,7)]

# find and remove values with zero variance 
        nzv <- nearZeroVar(training[,-55]+1, saveMetrics = T)
        training <- training[,nzv$zeroVar == FALSE]


# remove columns with NAs
# code at stack overflow re finding columns with missing values
        NaVal <- sapply(training, function (x) any(is.na(x) | x == ""))
        training <- training[,NaVal != T]


```

# Initial Analysis

Loading the training data into a tree from the tree package we are able to look at a plot of the data to see what trends may be evident. We can see that there are a number of leaves from the tree but can't see what may be causing them that at this point.  

The summary of the tree show the twelve variables most prominent in the creation of the tree.


```{r, cache=TRUE, echo=TRUE}
init_tree <- tree(classe~., data = na.omit(training))
 plot(init_tree)
summary(init_tree)


# find and remove values with high correlation
ncl <- ncol(training)
t <- findCorrelation(training[,-51], cutoff = 0.9, exact = F)
training <- training[,-t]


```


## Model Fitting  

I've chosen two model methods to fit over the data, being rpart to fit a regression model and random forest.  The train control option was selected as the initial random forest method was going over time.  

```{r, cache=TRUE, echo=TRUE}

trCont <- trainControl(method = "cv", number = 10 )
modFit <- train(classe ~., data = training, method ="rpart", trControl = trCont )
modFit2 <- train(classe ~., data = training, method ="rf", trControl = trCont)

```

The summary of the rpart model shows a median accuracy level of 37.9% which is not good enough to use for prediction.

```{r, cache=TRUE, echo = TRUE}
summary(modFit$results)
fancyRpartPlot(modFit$finalModel)
```
  
While the highest value of accuracy of the random forests is 85.3% as listed below.

```{r, cache=TRUE, echo = TRUE}
summary(modFit2$results)

```

The final answers for the quiz outcome can be obtained by predicting on modFit2 against the test data.
I'm not showing this as that isn't the point of this exercise.




